---
title: "diagFDR: mokapot diagnostics (competed winners)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{diagFDR: mokapot diagnostics (competed winners)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

This vignette demonstrates how to compute **diagFDR** diagnostics from **mokapot**
PSM exports.

A key choice is to compute diagnostics on **competed winners**: one PSM per
`run + spectrum_id`, selecting the maximum mokapot score among candidates
(targets and decoys). This corresponds to the standard *target–decoy competition*
setting for PSM-level inference.

The workflow is:

1. Read mokapot target and decoy PSM outputs.
2. Construct a competed-winner universe.
3. Run `dfdr_run_all()` and inspect tables/plots.
4. Optionally export results and a human-readable report.

## Runnable toy example (no external files required)

We start with a small simulated dataset that resembles a competed-winner mokapot output.
Any workflow producing a table that can be mapped to columns
`id`, `is_decoy`, `q`, `pep`, `run`, and `score` can similarly be used.

```{r toy}
library(diagFDR)

set.seed(2)

n <- 6000
toy <- data.frame(
  id = paste0("run1||scan", seq_len(n)),
  is_decoy = sample(c(FALSE, TRUE), n, replace = TRUE, prob = c(0.98, 0.02)),
  q = pmin(1, runif(n)^3),                                  # many small q-values
  pep = pmin(1, pmax(0, runif(n)^3 + rnorm(n, sd = 0.02))), # correlated toy PEP
  run = "run1",
  score = rnorm(n)
)

x <- as_dfdr_tbl(
  toy,
  unit = "psm",
  scope = "global",
  q_source = "toy mokapot",
  q_max_export = 1
)

diag <- dfdr_run_all(
  xs = list(mokapot = x),
  alpha_main = 0.01,
  alphas = c(5e-4, 1e-3, 2e-3, 5e-3, 1e-2, 2e-2, 5e-2, 1e-1, 2e-1),
  low_conf = c(0.2, 0.5)
)
```

### Headline stability at 1%

```{r headline}
diag$tables$headline
```

### Decoy tail support and stability proxy

```{r stability-plots}
diag$plots$dalpha
diag$plots$cv
```

### Local boundary support and elasticity

```{r boundary-stability}
diag$plots$dwin
diag$plots$elasticity
```

### Equal-chance plausibility by q-band (internal check)

```{r equal-chance}
diag$tables$equal_chance_pooled
diag$plots$equal_chance__mokapot
```

### PEP reliability and expected errors (ΣPEP)

```{r pep}
diag$tables$pep_IPE
diag$plots$pep_reliability__mokapot
```

```{r sumpep}
# sumpep is a list-of-tibbles (one per list)
diag$tables$sumpep$mokapot
```

## Real mokapot workflow (targets + decoys text exports)

The code below shows how to run **diagFDR** directly from mokapot outputs.

```{r real-mokapot, eval=FALSE}
library(diagFDR)

# Read mokapot outputs (targets + decoys)
raw <- read_mokapot_psms(
  target_path = "path/to/your_output.mokapot.psms.txt",
  decoy_path  = "path/to/your_output.mokapot.decoy.psms.txt"
)

# Construct competed winners (1 per run+spectrum_id; max mokapot score)
x <- mokapot_competed_universe(
  raw,
  id_mode = "runxid",        # recommended: id = run||spectrum_id
  unit = "psm",
  scope = "global",
  q_source = "mokapot q-value",
  q_max_export = 1
)

# Run diagnostics
diag <- dfdr_run_all(
  xs = list(mokapot = x),
  alpha_main = 0.01,
  alphas = c(5e-4, 1e-3, 2e-3, 5e-3, 1e-2, 2e-2, 5e-2, 1e-1, 2e-1),
  low_conf = c(0.2, 0.5)
)

# Write outputs to disk (tables + plots + summary + manifest + README)
dfdr_write_report(
  diag,
  out_dir = "diagFDR_mokapot_out",
  formats = c("csv", "png", "manifest", "readme", "summary")
)

# Render a single HTML report (requires rmarkdown in Suggests)
dfdr_render_report(diag, out_dir = "diagFDR_mokapot_out")
```

## Interpretation notes

- `D_alpha` (tail support) and `D_alpha_win` (local boundary support) quantify how well the
  cutoff is supported by decoys. Very strict thresholds can yield sparse decoy tails and thus
  unstable operating points (large `CV_hat`).

- `S_alpha(eps)` (elasticity) captures list sensitivity to small threshold perturbations.
  Low Jaccard overlap indicates that accepted IDs are sensitive to the chosen cutoff.

- Equal-chance diagnostics (decoy fraction in low-confidence q-bands) and PEP-based diagnostics
  (PEP reliability, ΣPEP) are **internal consistency checks** under target–decoy assumptions.
  They are informative but do not replace external validation strategies (e.g., entrapment) when
  decoy representativeness is uncertain.
